---
title: "postcodElapse - Usage guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{postcodElapse - Usage guide}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  
  )
library(tidyverse)
library(tidyterra)
theme_set(theme_minimal())
```

The goal of postcodElapse is to estimate air pollutant (PM<sub>2.6</sub>, BC, NO<sub>2</sub> & O<sub>3</sub>) concentration for Dutch postal codes. As a alternative to at location measurements. This is achieved trough getting geo-location data from one of two postcode databases. And mapping those locations to ELAPSE an air quality model for Europe. Estimating postcode-level air quality.

## Postcode databases
As said the geographic location or area is **required** to estimate air pollutant concentration. This data is expected as geopackage (.gpkg) databases which specialize in storing spacial data. postcodElapse support two geopackages:

1. [Postcode6](https://service.pdok.nl/cbs/postcode6/atom/postcode6_volledige_postcode.xml) by CBS mapping postcodes to geographic areas.
2. Kadaster's [Basisregistratie Adressen en Gebouwen(BAG)](https://service.pdok.nl/lv/bag/atom/bag.xml) containing the postcode and geo-location for every building in the Netherlands.

Enshure you downloat at-least one of threse.

## Effects of Low-Level Air Pollution: A Study in Europe
Or just ELAPSE is a Land Use Regression model. Estimating concentration of PM<sub>2.6</sub>, BC, NO<sub>2</sub> & O<sub>3</sub> air pollutants, across Europe at 100x100meter resolution. ELAPSE was published by [Kees de Hoogh](https://orcid.org/0000-0001-5974-2007) whom allowed it's use in postcodElapse. The model itself is stored as an Tiff grey-scale image with multiple layers:

- BCFULL
- NO2FULL
- O3FULLa
- O3FULLc
- O3FULLw
- PM25FULLt

ELAPSE contains three layers for O<sub>3</sub> due to it's prescience in the atmosphere being affected by the seasons. Typically in warmer(w) climate O<sub>3</sub> concentration tend higher. In the layer names *w* is warm, *c* is cold and *a* is the average.

The concentrations stored in ELAPSE all use the unit: µg/m<sup>3</sup>.

## postcodElapse
Now you have an idea of what postcodElapse does lets get some pollution estimates. For this demo I will assume you have downloaded both postcode database and postcodElapse itself.

The main function is `postcodElapse()` pass a postcode and database and it return pollution estimates. Ill be using BAG and the postcode of UMCG Gronningen. See the code below, yes it is that simple. It does take a minute, before you get your data.

```{r setup}
library(postcodElapse)

postcodElapse("9713AV", "../../data/bag-light.gpkg")
```

Oke the function tells you what it thinks the database is. It only recognizes BAG or PC6, you can manually override the guessing by passing a databases acronym in the *database_type* argument.

The pollution estimates are output as a data.frame. In the columns all layers of ELAPSE have been split across an average(avg), minimum(min) and maximum(max) pollution concentration, each row contains a single postcode. The *n* column is the amount of buildings found in that postcode.

Now lets run trough an 100 postcodes with PC6. You don't have to make up postcodes they are included as demo data. See the code below to load them.
```{r hunderd postcodes}
# The demo data is stored in extdata/postcod100.rda
path <- system.file("extdata/postcode100.rda", package = "postcodElapse")
load(path)

# Its an 100 postcodes common between BAG & PC6.
length(postcode100)
head(postcode100)

estimates <- postcodElapse(postcode100, "../../data/cbs_pc6_2024.gpkg") #Here I use PC6

head(estimates, 5)
```

Even tough I used PC6 the output is the same as BAG. If you want to test with more postcodes the file `postcodes1000.rda` is also present in extdata.

## Spacial plotting
The data postcodElapse handles inherently has an spacial component so plotting the data is of interest in some cases. There is one problem tough base R plotting doesn't handle spacial data well. I recommend you install the `tidyverse` and `tidyterra` packages, this equips us with ggplot2 and tidyterra enhances ggplot2 with spacial plotting.

### ELAPSE
Remember I said terra models air pollution concentration across Europe, well lets plot the internal one and took at the structure. See below.
```{r plotting ELAPSE}
# Remember to install these!
library(tidyverse)
library(tidyterra)

elapse <- loadElapse()

ggplot() +
  geom_spatraster(data = elapse) +
  facet_wrap(~ lyr)
```

