---
title: "postcodElapse - Usage guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{postcodElapse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The goal of postcodElapse is to estimate air pollutant (PM<sub>2.6</sub>, BC, NO<sub>2</sub> & O<sub>3</sub>) concentration for Dutch postal codes. As a alternative to at location measurements. This is achieved trough getting geo-location data from one of two postcode databases. And mapping those locations to ELAPSE an air quality model for Europe. Estimating postcode-level air quality.

## Postcode databases
As said the geographic location or area is **required** to estimate air pollutant concentration. This data is expected as geopackage (.gpkg) databases which specialize in storing spacial data. postcodElapse support two geopackages:

1. [Postcode6](https://service.pdok.nl/cbs/postcode6/atom/postcode6_volledige_postcode.xml) by CBS mapping postcodes to geographic areas.
2. Kadaster's [Basisregistratie Adressen en Gebouwen(BAG)](https://service.pdok.nl/lv/bag/atom/bag.xml) containing the postcode and geo-location for every building in the Netherlands.

Enshure you downloat at-least one of threse.

## Effects of Low-Level Air Pollution: A Study in Europe
Or just ELAPSE is a Land Use Regression model. Estimating concentration of PM<sub>2.6</sub>, BC, NO<sub>2</sub> & O<sub>3</sub> air pollutants, across Europe at 100x100meter resolution. ELAPSE was published by [Kees de Hoogh](https://orcid.org/0000-0001-5974-2007) whom allowed it's use in postcodElapse. The model itself is stored as an Tiff grey-scale image with multiple layers:

- BCFULL
- NO2FULL
- O3FULLa
- O3FULLc
- O3FULLw
- PM25FULLt

ELAPSE contains three layers for O<sub>3</sub> due to it's prescience in the atmosphere being affected by the seasons. Typically in warmer(w) climate O<sub>3</sub> concentration tend higher. In the layer names *w* is warm, *c* is cold and *a* is the average.

The concentrations stored in ELAPSE all use the unit: µg/m<sup>3</sup>.



```{r setup}
library(postcodElapse)
```
