---
title: "Spacial plotting"
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Put these here so we don't the output
library(tidyverse)
library(tidyterra)
```

# :construction: UNDER CONSTRUCTION :construction:
This article is incomplete!


Within this article the methods to plot the underlying spacial data postcodElapse work with will be explained. For this some preperation is required, we recommend installing the [tidyverse](https://tidyverse.org/), [tidyterra](https://dieghernan.github.io/tidyterra/articles/tidyterra.html) and [ggpubr](https://rpkgs.datanovia.com/ggpubr/) R-packages. These work together for spacial plotting in R.

## ELAPSE
To understand the spacial relation of postcodes some map is required, the best equivalent postcodElapse provides is ELAPSE. See the block below to load and plot ELAPSE.

```{r plotting ELAPSE}
library(tidyverse)
library(tidyterra)
library(postcodElapse)

elapse <- loadElapse()

ggplot() +
  geom_spatraster(data = elapse) +
  facet_wrap(~ lyr)
```

When plotting we recommend passing your data directly to the geom with the data argument. When no facet is added the average of all layers in ELAPSE will be plotted to a single figure. Use the `$` operator to specify which layer to plot. All concentrations are in the fill aesthetic with the unit: µg/m<sup>3</sup>. We recommend to assign the fill colour with: `?tidyterra::scale_fill_grass_c()`. 

## Accessing postcode geo-locations
To plot the location of postcodes on ELAPSE the geo-location is *required* these are provided trough: `pollutionFromBag()` and `pollutionFromPc6()`. These function output spacial data in the `geom` column. Plotting this data is achieved with `geom_sf()`, see the block below.

```{r layerd}
load(system.file("extdata/postcode100.rda", package = "postcodElapse"))

pc6 <- pollutionFromPc6(postcode100, "../../../data/cbs_pc6_2024.gpkg")
bag <- pollutionFromBag(postcode100, "../../../data/bag-light.gpkg")

ggplot() +
  geom_spatraster(data = elapse$NO2FULL) + 
  geom_sf(data = pc6, aes(geometry = geom), fill = "purple", alpha = 0.7) +
  geom_sf(data = bag, aes(geometry = geom), colour = "orange") +
  scale_fill_grass_c()
```

## Zooming and insetting
Due to the scale of ELAPSE in many situations it is hard to visualize the data. To alleviate this we recommend creating an inset plot and marking the location of the zoom on the main plot. postcodElapse provides helpers to do this see the block below.

```{r insetting}
library(ggpubr) #used to arrange plots into one figure.

main <- ggplot() +
  geom_spatraster(data = elapse$NO2FULL) + 
  geom_sf(data = pc6, aes(geometry = geom), fill = "purple", alpha = 0.7) +
  geom_sf(data = bag, aes(geometry = geom), colour = "orange", alpha = 0.7) +
  scale_fill_grass_c(transform = "log10") +
  geom_rectFeature(pc6[100, ], 5000) +
  labs(fill = bquote("log"[10]~" Concentration NO"[2]~"µg/m"^3)) +
  theme(legend.position = "top")

inset <- ggplot() +
  geom_spatraster(data = elapse$NO2FULL) + 
  geom_sf(data = pc6, aes(geometry = geom), fill = "purple", alpha = 0.7) +
  geom_sf(data = bag, aes(geometry = geom), colour = "orange", alpha = 0.7) +
  scale_fill_grass_c(transform = "log10") +
  coord_zoomFeature(pc6[100, ])

fig <- ggarrange(main, inset, ncol = 2,
                 common.legend = TRUE,
                 legend.grob = get_legend(main))

annotate_figure(fig, 
                fig.lab.pos = "bottom.left",
                fig.lab = "NO2 pollution in the Netherlands, with postcode location plotted. Orange is BAG purple is PC6.")
```

When using `geom_rectFeatrue()` and `coord_zoomFeature()` ensure you subset entire rows using `[12, ]`. Additionally air pollution estimate are always returned in alpha numeric order.
