---
title: "Spacial plotting"
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Put these here so we don't the output
library(tidyverse)
library(tidyterra)
```

Within this article the methods to plot the underlying spacial data
postcodElapse works with will be explored. For this some preparation is
required, we recommend installing the [tidyverse](https://tidyverse.org/),
[tidyterra](https://dieghernan.github.io/tidyterra/articles/tidyterra.html) and
[ggpubr](https://rpkgs.datanovia.com/ggpubr/) R-packages. These work together to
enable spacial plotting in R.

## ELAPSE
To understand the spacial relation of postcodes some map is required, the best
equivalent postcodElapse provides is ELAPSE. See the block below to load and
plot ELAPSE.

```{r plotting ELAPSE}
library(tidyverse)
library(tidyterra)
library(postcodElapse)

elapse <- loadElapse()

ggplot() +
  geom_spatraster(data = elapse) +
  facet_wrap(~ lyr)
```

When plotting we recommend passing the data directly to an geom via the *data* argument, this makes layering multiple datasets into one figure easy. The concentration retrieved from ELAPSE are mapped to the fill aestetic, to change the colour scheme we recommend `?tidyterra::scale_fill_grass_c()`. When no facet is used a single panel will be plotted containing mean pollutant concentrations. Subset ELAPSE with the `$` operator to plot a single layer.

## Accessing postcode geo-locations
To plot the location of postcodes on ELAPSE the geo-location is *required* these
are provided trough: `pollutionFromBag()` and `pollutionFromPc6()`. These
function output spacial data in the `geom` column. Plotting this data can be
done with `geom_sf()`, see the block below.

```{r layerd}
path <- system.file("extdata/postcode100.rds", package = "postcodElapse")
postcode100 <- readRDS(path)

pc6 <- pollutionFromPc6(postcode100, "../../../data/cbs_pc6_2024.gpkg")
bag <- pollutionFromBag(postcode100, "../../../data/bag-light.gpkg")

ggplot() +
  geom_spatraster(data = elapse$NO2FULL) + 
  geom_sf(data = pc6, aes(geometry = geom), fill = "purple", alpha = 0.7) +
  geom_sf(data = bag, aes(geometry = geom), colour = "orange") +
  scale_fill_grass_c()
```

## Zooming and insetting
Due to the scale of ELAPSE it may be hard to visualize the data. To alleviate this we recommend creating an inset plot and marking the location of the zoom on the main plot. postcodElapse provides `geom_rectFeature()` and `coord_zoomFeature()` to annotate and zoom into specific spatial features. Below is an example showing how to use these helpers with ggpubr to build an inset.

```{r insetting}
library(ggpubr) #used to arrange plots into one figure.

main <- ggplot() +
  geom_spatraster(data = elapse$NO2FULL) + 
  geom_sf(data = pc6, aes(geometry = geom), fill = "purple", alpha = 0.7) +
  geom_sf(data = bag, aes(geometry = geom), colour = "orange", alpha = 0.7) +
  scale_fill_grass_c(transform = "log10") +
  geom_rectFeature(pc6[100, ], 5000) + # Mark the location of interest
  labs(fill = bquote("log"[10]~" Concentration NO"[2]~"µg/m"^3)) +
  theme(legend.position = "top")

inset <- ggplot() +
  geom_spatraster(data = elapse$NO2FULL) + 
  geom_sf(data = pc6, aes(geometry = geom), fill = "purple", alpha = 0.7) +
  geom_sf(data = bag, aes(geometry = geom), colour = "orange", alpha = 0.7) +
  scale_fill_grass_c(transform = "log10") +
  coord_zoomFeature(pc6[100, ]) # Zoom into the location of interest

# Make and annotate the combined figure.
fig <- ggarrange(main, inset, ncol = 2,
                 common.legend = TRUE,
                 legend.grob = get_legend(main))

annotate_figure(fig, 
                fig.lab.pos = "bottom.left",
                fig.lab = "NO2 pollution in the Netherlands, with postcode location plotted. Orange is BAG purple is PC6.")
```

When using `geom_rectFeatrue()` and `coord_zoomFeature()` ensure you subset entire rows using `[12, ]`. Additionally air pollution estimate are always returned in alpha numeric order.

With this material you can plot the spacial data of postcodes and ELAPSE. To see all plotting options we recommend the documentation of  [ggplot](https://ggplot2.tidyverse.org/), [tidyterra](https://dieghernan.github.io/tidyterra/articles/tidyterra.html) and [ggpubr](https://rpkgs.datanovia.com/ggpubr/index.html_)
