---
title: "Spacial plotting"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(postcodElapse)
```

# :construction: UNDER CONSTRUCTION :construction:
This article is incomplete!

## Spacial plotting
The data postcodElapse handles inherently has an spacial component so plotting the data is of interest in some cases. There is one problem tough base R plotting doesn't handle spacial data well. I recommend you install the `tidyverse` and `tidyterra` packages, this equips us with ggplot2 and tidyterra enhances ggplot2 with spacial plotting.

### ELAPSE
Remember I said terra models air pollution concentration across Europe, well lets plot the internal one and took at the structure. See below.
```{r plotting ELAPSE}
# Remember to install these!
library(tidyverse)
library(tidyterra)

elapse <- loadElapse()

ggplot() +
  geom_spatraster(data = elapse) +
  facet_wrap(~ lyr)
```

We can see the shape of the Netherlands well, if you want to change the fill color I recommend reading the docs of `?tidyterra::scale_fill_grass_c`. Now we are going to plot the location of postcodes ontop of ELAPSE.

For this we are using `pollutionFromPc6()` which is one layer deeper into postcodeLapse, first lets run it without plotting

```{r layerd plotting}
data_pc6 <- pollutionFromPc6("9713AV", "../../../data/cbs_pc6_2024.gpkg")

data_pc6
```
