---
title: "Spacial plotting"
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
# Put these here so we don't the output
library(tidyverse)
library(tidyterra)
```

This article describes the method to plot the spacial data underlying postcodElapse. To enable spatial plotting in R, installation of of the R packages [tidyverse](https://tidyverse.org/),
[tidyterra](https://dieghernan.github.io/tidyterra/articles/tidyterra.html) and
[ggpubr](https://rpkgs.datanovia.com/ggpubr/) is required.

## ELAPSE
First, a map with the spatial relations of the postal codes is required. In postcodElapse, ELAPSE is the best equivalent, which can be loaded and plotted as indicated below: 

```{r plotting ELAPSE}
library(tidyverse)
library(tidyterra)
library(ggpubr)
library(postcodElapse)

elapse <- loadElapse()

ggplot() +
  geom_spatraster(data = elapse) +
  facet_wrap(~ lyr)
```

When plotting spatially pass the data for a layer directly into a geom via the *data* argument, plotting a single layer is done by sub-setting ELAPSE with `$`. The `tidyterra::geom_spatraster()` map air pollutant levels to the fill aesthetic, use `tidyterra::scale_fill_grass_c()` to alter the fill colour.

# Accessing postal coded geo-locations
To plot the location of postcodes on ELAPSE the geo-location is *required* these
are provided trough: `pollutionFromBag()` and `pollutionFromPc6()`. These
function output spacial data in the `geom` column. Plotting this data can be
done with `geom_sf()`, see the block below.

```{r layerd}
path <- system.file("extdata/postcode100.rds", package = "postcodElapse")
postcode100 <- readRDS(path)

pc6 <- pollutionFromPc6(postcode100, "../../../data/cbs_pc6_2024.gpkg")
bag <- pollutionFromBag(postcode100, "../../../data/bag-light.gpkg")

ggplot() +
  geom_spatraster(data = elapse$NO2FULL) + 
  geom_sf(data = pc6, aes(geometry = geom), fill = "purple", alpha = 0.7) +
  geom_sf(data = bag, aes(geometry = geom), colour = "orange") +
  scale_fill_grass_c() +
  labs(fill = bquote("NO"[2]~"concentration in µg/m"^3))
```

## Zooming and insetting
The scale of ELAPSE prohibits the visualization of the data. To alleviate this, it is recommended to create an inset plot and mark the location of the zoom in the main plot. postcodElapse provides `geom_rectFeature()` and `coord_zoomFeature()` to annotate and zoom into specific spatial features. An example is proved below.

```{r insetting}
# Build the plots for the single figure separately
main <- ggplot() +
  geom_spatraster(data = elapse$NO2FULL) + 
  geom_sf(data = pc6, aes(geometry = geom), alpha = 0.7, fill = "purple") +
  geom_sf(data = bag, aes(geometry = geom), alpha = 0.7, colour = "orange") +
  scale_fill_grass_c(transform = "log10") +
  geom_rectFeature(pc6[100, ], 5000) + # Annotate a postal address
  labs(fill = bquote("log10 Concentration NO"[2]~"in µg/m"^3)) +
  theme(legend.position = "top")

inset <- ggplot() +
  geom_spatraster(data = elapse$NO2FULL) +
  geom_sf(data = pc6, aes(geometry = geom), alpha = 0.7, fill = "purple") +
  geom_sf(data = bag, aes(geometry = geom), alpha = 0.7, colour = "orange") +
  scale_fill_grass_c(transform = "log10") +
  coord_zoomFeature(pc6[100, ]) # Zoom into a postal address

# Combine the two plots into a single figure.
fig <- ggarrange(main, inset,
  ncol = 2,
  common.legend = TRUE,
  legend.grob = get_legend(main)
)

# Lable the figure
label <- 
"NO2 pollutant levels at postal code level in the Netherlands.
Purple areas are postal adrreses from PC6, orange dots are buildings from BAG."

annotate_figure(fig, 
  fig.lab.pos = "bottom.left",
  fig.lab = label
)
```

When using `geom_rectFeatrue()` and `coord_zoomFeature()` ensure you subset entire rows using `[12, ]`. Additionally air pollution estimate are always returned in alpha numeric order.

With this material you can plot the spacial data of postcodes and ELAPSE. To see all plotting options we recommend the documentation of  [ggplot](https://ggplot2.tidyverse.org/), [tidyterra](https://dieghernan.github.io/tidyterra/articles/tidyterra.html) and [ggpubr](https://rpkgs.datanovia.com/ggpubr/index.html_)
