---
title: "Spacial plotting"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tidyverse)
library(tidyterra)
library(postcodElapse)
```

# :construction: UNDER CONSTRUCTION :construction:
This article is incomplete!

## Spacial plotting
The data postcodElapse handles inherently has an spacial component so plotting the data is of interest in some cases. There is one problem tough base R plotting doesn't handle spacial data well. I recommend you install the `tidyverse` and `tidyterra` packages, this equips us with ggplot2 and tidyterra enhances ggplot2 with spacial plotting.

### ELAPSE
Remember I said terra models air pollution concentration across Europe, well lets plot the internal one and took at the structure. See below.
```{r plotting ELAPSE}
# Remember to install these!
library(tidyverse)
library(tidyterra)

elapse <- loadElapse()

ggplot() +
  geom_spatraster(data = elapse) +
  facet_wrap(~ lyr)
```

We can see the shape of the Netherlands well, if you want to change the fill colour I recommend reading the docs of `?tidyterra::scale_fill_grass_c`. Before we plot the location of postcodes on top of ELAPSE, I am going to explain the `pollutionFrom*()` functions.

### PollutionFrom*()
The `pollutionFrom*()` functions do the estimating of air-quality for postcodes, they are of course specific to BAG or PC6. More importantly they output spacial data, `postcodElapse()` removes that. First lets look at the spacial data from PC6.

```{r raw data PC6}
pollutionFromPc6("9713AV", "../../../data/cbs_pc6_2024.gpkg")

```

As you see it looks like the output from `postcodElapse()` but it has extra data. All the `Simple features collecto...` stuff is metadata describing the coordinate system, you don't have to worry about this unless you need to modify it. In the data itself the extra column **geom** exists, it contains the physical area of an postcode. Now lets look at BAG

```{r raw data BAG}
data <- pollutionFromBag("9714AV", "../../../data/bag-light.gpkg")

head(data, 4)
```

BAG has way less columns than PC6 that is due to bag containing points. It is not practical to convert points to areas. So I built `pollutionFromBag()` to output the raw data collected from BAG and ELAPSE. It does provide a-lot more rows one per building in a postcodes for all postcodes. The geom column contains effectively the geo-graphic position of a building.

## Plotting
Now we understand the `pollutionFrom*()` functions we do some spacial plotting with layers. I have selected some postcodes form public locations across the Netherlands and we are going to plot then on ELAPSE.

```{r basic layers}
postcodes <- c("9714AV", "1071XX", "8934CJ", "1789AX")

elapse <- loadElapse()

pc6 <- pollutionFromPc6(postcodes, "../../../data/cbs_pc6_2024.gpkg")
bag <- pollutionFromBag(postcodes, "../../../data/bag-light.gpkg")

ggplot() +
  geom_spatraster(data = elapse$NO2FULL) +
  geom_sf(data = pc6, aes(geometry = geom), fill = "purple", alpha = 0.7) +
  geom_sf(data = bag, aes(geometry = geom), colour = "orange") +
  scale_fill_grass_c()
```

Oke we can see ELAPSE and see orange points, where are the purple areas? And shouldn't there be many points? They are both in the plot just very small and fighting over the same space. Lets zoom in on the Rijskmuseum the 1st postcode.
Postcodes are always output in alpha numecic fastion!

```{r Zooming}
postcodes <- c("9714AV", "1071XX", "8934CJ", "1789AX")

elapse <- loadElapse()

pc6 <- pollutionFromPc6(postcodes, "../../../data/cbs_pc6_2024.gpkg")
bag <- pollutionFromBag(postcodes, "../../../data/bag-light.gpkg")

ggplot() +
  geom_spatraster(data = elapse$NO2FULL) +
  geom_sf(data = pc6, aes(geometry = geom), fill = "purple", alpha = 0.7) +
  geom_sf(data = bag, aes(geometry = geom), colour = "orange") +
  coord_zoomFeature(pc6[1]) +
  scale_fill_grass_c()
```
